---
title: "pcadapt - combined deceased/survivor run"
output: html_notebook
---

This uses the already-built LD-thinned, un-split dataset (`merged_LD_thinned`)
so pcadapt sees both groups together. We then check which PC (if any)
separates deceased from survivors, and run a componentwise outlier test on
that specific axis -- rather than pcadapt's default Mahalanobis distance,
which combines all K PCs and would mix "survival-related" structure with
any other structure (batch, family, geography) picked up by the other PCs.

No new plink step is needed -- `merged_LD_thinned.bed/.bim/.fam` from your
original pipeline is exactly what we want here (it's the LD-pruned dataset
containing both groups, before the --keep split).

```{r}
library(pcadapt)
library(dplyr)
library(readxl)
library(tidyr)
library(fuzzyjoin)
```

## Gene coordinate table

One table, one place to maintain coordinates -- this replaces the repeated
per-chromosome `case_when` blocks used earlier, which is what let a
chromosome typo (`SUPER_7` vs `SUPER_1`) and a mislabeled range (TLR7 vs
TLR15) slip through undetected. With everything in a single table, a typo
here breaks obviously (e.g. an empty result for that gene) rather than
silently mislabeling another gene's hits.

```{r}
genes <- data.frame(
  gene   = c("TLR1A", "TLR1B", "TLR2A", "TLR2B", "TLR3-1", "TLR3-2", "TLR3-3", "TLR3-4",
             "TLR4-1", "TLR4-2", "TLR4-3", "TLR5", "TLR7", "TLR15", "TLR21"),
  region = c("SUPER_4:53881681-53884137",
             "SUPER_4:53873407-53875357",
             "SUPER_4:32245726-32248107",
             "SUPER_4:32236631-32238982",   # corrected: was previously swapped with TLR3-1
             "SUPER_4:20104648-20105088",   # corrected: was previously swapped with TLR2B
             "SUPER_4:20103611-20103802",
             "SUPER_4:20100233-20102059",
             "SUPER_4:20099425-20099652",
             "SUPER_18:8042591-8042695",
             "SUPER_18:8041589-8041755",
             "SUPER_18:8037993-8040252",
             "SUPER_3:95457095-95459677",
             "SUPER_1:77572878-77576018",
             "SUPER_3:116327465-116330056",
             "SUPER_29:1579922-1582852")
) %>%
  separate(region, into = c("chr", "range"), sep = ":") %>%
  separate(range, into = c("start", "end"), sep = "-", convert = TRUE)

genes
```

## 1. Read the genotype data

```{r}
path_to_file <- "C:/Users/shayl/OneDrive - Queen's University/TLRPaper/bedfiles/merged_LD_thinned.bed"
merged_data <- read.pcadapt(path_to_file, type = "bed")
```

## 2. Build a group label vector, in the same sample order as the .fam file

pcadapt doesn't carry sample IDs through internally, so the group vector
you build here MUST be in the exact same row order as the .fam file used
above, or every downstream label will be wrong.

```{r}
fam_path <- "C:/Users/shayl/OneDrive - Queen's University/TLRPaper/bedfiles/merged_LD_thinned.fam"
fam <- read.table(fam_path, stringsAsFactors = FALSE)
colnames(fam)[1:2] <- c("FID", "IID")

meta <- read_excel("sample_metadata.xlsx", sheet = "depth_summary")

# the .fam IDs have the trailing "_trimmed_sorted_minq20" stripped relative
# to the metadata IDs (same issue as the --keep files earlier) -- strip it
# here too so the join matches
meta <- meta %>%
  mutate(IID = sub("_trimmed_sorted_minq20$", "", IID))

# join in fam order (do NOT sort/reorder -- must match the bed file row order)
poplist <- fam %>%
  left_join(meta, by = "IID") %>%
  pull(group)

# sanity checks before trusting anything downstream
stopifnot(nrow(fam) == length(poplist))
sum(is.na(poplist))          # should be 0 -- any NA means an unmatched sample
table(poplist)               # should read ~39 deceased / ~31 survivor
```

If `sum(is.na(poplist))` is not 0, stop here and fix the ID matching before
going further -- an NA in `poplist` means that sample's group is unknown and
everything downstream (coloring, PC-group tests, outlier calls) will be
misaligned.

## 3. Choose K

```{r}
x <- pcadapt(input = merged_data, K = 20)
plot(x, option = "screeplot")
```

Pick K just past the "elbow" as before. For a two-group comparison you want
enough PCs to have a decent chance of capturing a survival-related axis, but
not so many that you're just fitting noise -- K = 4-6 is a reasonable start;
adjust based on the screeplot.

```{r}
K_final <- 4   # <-- set based on your screeplot
y <- pcadapt(input = merged_data, K = K_final, method = "componentwise")
summary(y)
```

## 4. Find which PC (if any) separates deceased from survivors

`get.pc()` isn't needed here -- we can pull PC scores directly from the
pcadapt object (`y$scores`) and test each one against group.

```{r}
scores <- as.data.frame(y$scores)
colnames(scores) <- paste0("PC", 1:K_final)
scores$group <- poplist

# quick visual check
plot(x, option = "scores", pop = poplist)
plot(x, i = 3, j = 4, option = "scores", pop = poplist)

# formal check: does each PC differ by group?
for (i in 1:K_final) {
  pc_col <- paste0("PC", i)
  test <- wilcox.test(scores[[pc_col]] ~ scores$group)
  cat(pc_col, "- p =", test$p.value, "\n")
}
```

Look for the PC(s) with a small p-value here -- that's your candidate
survival-associated axis. If none of the PCs separate the groups at all,
that itself is a real (if less exciting) result: it suggests deceased and
survivors aren't strongly differentiated on the axes pcadapt is picking up,
and a global genome scan may not be the most sensitive way to detect
selection here (an Fst-based scan targeted at your TLR windows specifically
may still be worth doing regardless).

## 5. Componentwise outlier test on the survival-associated PC

Suppose PC2 came out significant above -- adjust `survival_pc` to whichever
PC number was actually significant in your output.

```{r}
survival_pc <- 2   # <-- set this based on step 4's results

pvals_pc <- y$pvalues[, survival_pc]

library(qvalue)
qval <- qvalue(pvals_pc)$qvalues
alpha <- 0.05
merged_outliers <- which(qval < alpha)
length(merged_outliers)

merged_outliers <- as.data.frame(merged_outliers)
write.csv(merged_outliers, file = "merged_outliers_componentwise.csv")
```

## 6. Match outliers back to variant positions

Same rule as before: match against the .bim file that was actually used in
this pcadapt run (`merged_LD_thinned.bim`), not any other bed/bim file.

```{bash}
cut -d',' -f2 merged_outliers_componentwise.csv > merged_outliers.txt
sed -i '1d' merged_outliers.txt
dos2unix merged_outliers.txt

awk 'NR==FNR{line[$1]; next} FNR in line' merged_outliers.txt merged_LD_thinned.bim > merged_outliers_pcadapt.txt
```

## 7. Filter to TLR positions of interest

Instead of one hand-written filter block per chromosome, this does a single
range join against the `genes` table above: any outlier SNP whose chr+pos
falls inside a gene's start-end window gets that gene's label attached.
Adding, correcting, or removing a gene now only requires editing the
`genes` table -- there's no separate filtering logic to keep in sync with it.

```{r}
merged_hits <- read.table("merged_outliers_pcadapt.txt") %>%
  select(c(1, 4, 5, 6)) %>%
  rename(chr = V1, pos = V4, base_before = V5, base_after = V6) %>%
  mutate(start = pos, end = pos)   # single-position "range" so genome_join can compare it to gene windows

combined_merged_outliers <- genome_left_join(
  merged_hits, genes,
  by = c("chr", "start", "end")
) %>%
  filter(!is.na(gene)) %>%          # keep only SNPs that actually fall inside a TLR window
  select(chr = chr.x, pos, base_before, base_after, gene)

write.csv(combined_merged_outliers, file = "pcadapt_merged_tlr_outliers.csv")
combined_merged_outliers
```

Sanity check worth running once: confirm every gene in the `genes` table
that you expect to see hits for is at least *capable* of appearing (i.e. no
gene silently vanishes because of a typo in its `region` string):

```{r}
genes$gene[!genes$gene %in% combined_merged_outliers$gene]   # genes with zero outliers in this run -- expected for most, just eyeball for anything suspicious
```
